---
title: "Exercises from Week 2, Lecture 2: Subsetting"
subtitle: "SDS 270"
author: "Jericho Lawson"
format: html
---

## Introduction

In these exercises, we will be working with `starwars` tibble to learn how to subset in various manners.

By the end of this lab, you will be able to:

- subset a data frame and tibble based on positive, negative, logical, and character indexing
- apply the `filter` and `select` functions via `dplyr`

Load in the following libraries below to work on this lab: `tidyverse`

```{r}
library(tidyverse)
```

You will create a separate .qmd file and complete the following exercises that will then be committed and pushed in your SDS 270 repository.


## Exercises

For these exercises, you will be working exclusively with the `starwars` tibble. Answer the following questions below:

::: {.callout-important}

##### Q1

Using a `dplyr` function, construct a tibble from `starwars` that only contains characters with blue eyes.

```{r}
#| output: false
# Do not alter code right above.

```

:::

::: {.callout-important}

##### Q2

Using a `dplyr` function, construct a tibble from `starwars` that only contains characters with blue eyes and a height below 100 cm. Then, display only the names of these characters.

```{r}
#| output: false
# Do not alter code right above.

```

:::

### Base R Subsetting vs. dplyr() functions

In many applications, you will find R users that subset through the `[]`, `[[]]`, `$` symbols, but unless we know the type and functionality associated with the type, we will run across some errors.

The next few exercises will showcase this. You will use the following objects in the next 3 exercises.

```{r}
num <- 5
var <- "skin_color"
```


::: {.callout-important}

##### Q3

Using `starwars`, access the 5th column using the `num` object within the `[[]]`, `[]` and `$` operators. What do you notice?

```{r}
#| output: false
#| eval: false
# Do not alter code right above.

```


:::


::: {.callout-important}

##### Q4

Using `starwars`, access the 5th column using the `var` object within the `[[]]`, `[]` and `$` operators. What do you notice?

```{r}
#| output: false
#| eval: false
# Do not alter code right above.

```


:::

::: {.callout-important}

##### Q5

Using `starwars`, access the 5th column by name with one of the `dplyr` functions. What do you notice?

```{r}
#| output: false
#| eval: false
# Do not alter code right above.

```


:::


### List-columns: `nest()` and `unnest()`

We now look into list-columns, which are columns in which each element is composed of a list. When we think of a vector, most of the time we see a single value that is stored in an element. A list-column changes this. That value is now a list, which can be of any length we desire. 

List-columns can be expanded by `unnest()`. This has the effect of lengthening the data frame (sort of like an accordion). Each row is duplicated for each unique value of each entry in the list-column.

The `nest()` function performs the opposite operation of "rolling up" the data frame to create a new list-column. We did this in the intro of list-columns.

::: {.callout-important}

##### Q6

Now using the `starwars` tibble, experiment with list-columns by expanding and contracting the other list-columns in the storms data frame.

:::

### Mapping over list columns
Suppose now we want to add the numbers of films for each character to the `starwars` data set. A simple `mutate()` like this will not throw an error, but also won’t do what we want.

```{r}
oops <- starwars |>
  mutate(num_films = length(films)) |>
  arrange(desc(num_films)) |>
  select(name, num_films)
oops
```

This just made all of the entries equal to `length(films)`.

```{r}
all(oops$num_films == length(starwars$films))
```

To get this right, we need to `map()` inside our `mutate()`.

```{r}
starwars |>
  mutate(num_films_actual = map_int(films, length)) |>
  arrange(desc(num_films_actual)) |>
  select(name, num_films_actual, films)
```

