---
title: "Exercises from Week 4, Lecture 2: Functionals"
subtitle: "SDS 270"
format: html
---

## Introduction

In these exercises, we will explore particular examples where the use of a functional improves the efficiency and interpretability of the code.

```{r}
#| echo: false
#| warning: false
library(tidyverse)
```

## Combing through Datasets

Suppose we are looking through the following three datasets (all as objects in base R):

- `mtcars`
- `iris`
- `trees`

In these data sets, we will gather some basic information.

::: {.callout-important}

### Q1

Create an object named `datasets` that has an appropriate data structure that can house all three datasets specified. 

```{r}
# PUT SOLUTION HERE
```

:::

### Using a For Loop

::: {.callout-important}

### Q2

Create an empty vector of length 3 that will store the number of observations in each of the datasets. From there, write a for-loop that gathers the number of observations in each dataset, storing that number inside the created vector.

Write this in between the `Sys.time` code.

```{r}
begin_for <- Sys.time()

# PUT SOLUTION HERE

time_for <- Sys.time() - begin_for
```

:::

### Using `apply()`

::: {.callout-important}

### Q3

Redo Q2 by using an appropriate `apply()` statement instead.

```{r}
begin_apply <- Sys.time()

# PUT SOLUTION HERE

time_apply <- Sys.time() - begin_apply
```

:::

### Comparing the Two Methods

::: {.callout-important}

### Q4

Compare the two methods with regard to efficiency and code. Which method is faster? What are some pros and cons of using either method?

PUT SOLUTION HERE

:::

## Metadata for the User

Let's expand on the ideas seen previously by extracting several tidbits of information from the datasets to print to the user. 

::: {.callout-important}

### Q5

Come up with four different pieces of information about a dataset that may be useful to an analyst at first glance.

PUT SOLUTION HERE

:::

::: {.callout-important}

### Q6

Using what you typed up for Q5, create a function that takes in some dataset as `x` and returns all aforementioned information in an appropriate data structure.

```{r}
# PUT SOLUTION HERE
```


:::

::: {.callout-important}

### Q7

Write an appropriate `apply()` statement that uses your function from Q6 on each dataset in your `datasets` object.

```{r}
# PUT SOLUTION HERE
```
:::

## Turn-In

Stage, commit, and push your template file to your SDS 270 repository.
