---
title: "Exercises from Week 5, Lecture 1: Functionals - `map()`"
subtitle: "SDS 270"
author: "NAME HERE"
format: html
---

## Introduction

We will utilize CO2 concentration data from Mauna Loa between 1959-1997. More information about the dataset can be found by using `?co2`. Run the following code below to create a data frame with that `co2` data:

```{r, message = FALSE, warning = FALSE}
library(purrr)
co2_df <- as.data.frame(matrix(co2, nrow = 39, ncol = 12, byrow = T))
rownames(co2_df) <- 1959:1997
colnames(co2_df) <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
                   "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
```

With this data frame, we will find out how we can use the various `map()` functions to complete particular analytical tasks. 

::: {.callout-important}

### Q1

Using the `co2_df` dataset, subset the data into two different datasets, with the following specifications:

- `co2_60s`: Data from 1960-1969
- `co2_70s`: Data from 1970-1979

```{r}
# PUT SOLUTION HERE
```

:::

::: {.callout-important}

### Q2

Generate a summary of every single month (column) from the `co2_60s` data. 

```{r}
# PUT SOLUTION HERE
```

:::

::: {.callout-important}

### Q3

Create a function that takes the mean of the difference of two atomic vectors (as inputs `x`, `y`) and returns this mean. Then, use one of the `map()` functions to apply the created function on the `co2_60s` and `co2_70s`months.


```{r}
# PUT SOLUTION HERE
```

:::

::: {.callout-important}

### Q4

Find the `minimum` monthly CO2 level across the years in the `co2_60s` data.

```{r}
# PUT SOLUTION HERE
```

:::


## Turn-In

Stage, commit, and push your template file to your SDS 270 repository.
