---
title: "Exercises from Week 7, Lecture 1: Conditions"
subtitle: "SDS 270"
autor: "YOUR NAME HERE"
format: html
---

## Introduction

In these exercises, we will learn how to handle conditions.

**Goal:** By the end of this lab, you should be able to handle an error and return from it gracefully.

```{r}
library(tidyverse)
```


## Three types of conditions

Recall that there are three main types of conditions in R:

- **error**: will terminate the function immediately and throw an error
- **warning**: will continue to execute the function, but will warn you about what happened afterwards
- **message**: will immediately display a message about what is happening.

The following function will compute the number of unique values in whatever columns in a data frame you select. The first argument is the data frame, and the second argument is the dots, which are simply passed to `select()`.

Note that this function may display a message, a `warning()`, or an error depending on what happens.

```{r}
unique_values <- function(.data, ...) {
  x <- .data |>
    select(...)
  
  if (ncol(x) == ncol(.data)) {
    warning("All columns selected!")
  }
  
  if (ncol(x) < 1) {
    stop("No columns selected.")
  }

  message(paste("Computing on", ncol(x), "columns..."))  
  map_int(x, n_distinct)
}
```

::: {.callout-important}

### Q1

Examine the code for `unique_values()` carefully. Explain how the function works.

PUT SOLUTION HERE

:::

If we give this function no valid selection, we will throw an error.

```{r}
#| error: true
unique_values(starwars)
```

::: {.callout-important}

### Q2

Was the error in the previous example thrown by `unique_values()` or by `select()`? How do you know?

PUT SOLUTION HERE

:::

If we select all of the columns in the data frame, then we get a warning. However, the code still executes.

```{r}
#| error: true
unique_values(starwars, everything())
```

::: {.callout-important}

### Q3

Why does the warning message show up before the results? Why does the warning message show up before the message?

PUT SOLUTION HERE

:::

Since `unique_values()` passes the dots to `select()`, we can leverage all of the functionality of the select helpers!

```{r}
#| error: true
unique_values(starwars, contains("n"))
```

However, if we pass garbage to `select()`, then of course `select()` will still throw an error.

```{r}
#| error: true
unique_values(starwars, i_love_r)
```

::: {.callout-important}

### Q4

Consider the difference in the output between the previous example and `unique_values(starwars)`. What is different?

PUT SOLUTION HERE

:::

## Catching errors

Instead of just failing whenever the user passes bad arguments to `select()`, we might want to catch those errors and do something with them. Here, we use a `tryCatch()` statement to provide some additional information about what went wrong, and to continue with the original data frame if the `select()` statement failed.

```{r}
unique_values_safe <- function(.data, ...) {
  x <- tryCatch(
    error = function(cnd) {
      warning("Attempt to select column has failed")
      message("Here is what we know about the error")
      str(cnd)
      .data
    },
    .data |>
      select(...)
  )
  
  if (ncol(x) == ncol(.data)) {
    warning("All columns selected!")
  }
  
  if (ncol(x) < 1) {
    stop("No columns selected.")
  }

  message(paste("Computing on", ncol(x), "columns..."))  
  map_int(x, n_distinct)
}
```

Now, even though an error still occurs, we still get output.

```{r}
#| error: true
unique_values_safe(starwars, i_love_r)
```

Whether this output is sensible is an open question for the developer. In this case I think it is probably not sensible.

::: {.callout-important}

### Q5

Why might it be a better idea to fail with an error in the previous example instead of continuing with the full data frame?

PUT SOLUTION HERE

:::


## Turn-In

Stage, commit, and push your template file to your SDS 270 repository.
