---
title: "Exercises from Week 7, Lecture 2: Object-Oriented Programming"
subtitle: "SDS 270"
autor: "YOUR NAME HERE"
format: html
---
## Introduction

In these exercises, we will learn how to write methods for S3 generics.

Goal: by the end of this lab, you should be able to...

- write your own S3 generic methods and functions, and
- identify pieces that make up object-oriented programming.

(Install and) load in the following libraries below to work on this lab: `sloop`, `tidyverse`, `emoji`, and `crayon`

```{r}
#| warning: false
library(sloop)
library(tidyverse)
library(emoji)
library(crayon)
```

## Exercises: Set I

These exercises will be completed during lecture.

In groups of 2-3, think of two (real-world) objects that are adjacent to a car in some fashion. Consider the following:

::: {.callout-important}

##### Q1

- For each (real-world) object, what attributes and functions does this object have?

PUT ANSWER HERE

:::

::: {.callout-important}

##### Q2

- For one function for each object, what is needed to start and end the function?

PUT ANSWER HERE

:::


::: {.callout-important}

##### Q3

- What connects the (real-world) objects together with a car object?

PUT ANSWER HERE

:::

## Exercises: Set II

There are many methods for the `print()` generic available:

```{r}
s3_methods_generic("print")
```

Other common S3 generics include `summary()` and `plot()`:

```{r}
s3_methods_generic("summary")
s3_methods_generic("plot")
```


::: {.callout-important}

##### Q4

Try to think of another generic function that you’ve worked with in the past. Use `s3_methods_generic()` to find its available methods. [Note that you might have to load other packages to find the methods or the generic that you recall.]

```{r}
# ANSWER HERE
```

:::

In addition to listing the methods for a particular generic, we can find all the methods that are defined for a particular class using `s3_methods_class()`:

```{r}
s3_methods_class("tbl_df")
```

Note that while the methods for a particular **generic** are most likely written by many different developers in many different packages, the **methods** for a particular class are most likely written by the same set of developers of a single package.

### Method dispatch

If you are ever unsure of which method is actually being run when you invoke a generic, use `s3_dispatch()` to figure out what is going on.

```{r}
s3_dispatch(print(starwars))
s3_dispatch(print(mtcars))
```


::: {.callout-important}

##### Q5

Why do the previous two examples show a different list of available methods?

PUT ANSWER HERE

:::

::: {.callout-important}

##### Q6

Why doesn’t `print.data.frame()` get called on `starwars`?

PUT ANSWER HERE

:::

### Using a method

First, we’ll learn about two fun packages: `emoji` and `crayon`.

#### `emoji`
The [`emoji`](https://emilhvitfeldt.github.io/emoji/) package allows you print an emoji via the `emoji()` function. To display an emoji, call it by name:

```{r}
emoji("star")
emoji("dog")
emoji("basketball")
```

You can see the full list of available emojis in the tibble returned by emojis.

- `emojis |> View()`

::: {.callout-important}

##### Q7

Find an emoji that you like and use emoji() to print it here.

```{r}
# PUT ANSWER HERE
```

:::


#### `crayon`
The [`crayon`](https://github.com/r-lib/crayon) package allows you to print in color in the console. For a complicated example of what you can do, consider the tidyverse logo:

```{r}
tidyverse_logo()
```

For a simpler, example, you can write your name in blue. Be sure to wrap the text in `cat()` (or `message()`) to send it to the console.

```{r}
cat(crayon::blue("Jericho"))
```

::: {.callout-important}

##### Q8

Pick a color from the crayon package and write something using that color.

```{r}
# PUT ANSWER HERE
```

:::

### Defining a method
Now we will define a simple `print()` method that will only be invoked for objects of class `sds270`. This method will simply print a colorful message to the console, and then invoke the `NextMethod()` available.

```{r}
print.sds270 <- function(x, ...) {
  cat(
    crayon::magenta(
      "Jericho", 
      emoji::emoji("heart"), 
      "s programming in R!\n"
    )
  )
  NextMethod()
}
```


To use it, we just have to create an object of type `sds270`, and then call print(). Remember that in R you can simply modify the class attribute of any object!

```{r}
class(starwars) <- c("sds270", class(starwars))
starwars
```

Note that this method works on objects of any underlying class!!

```{r}
class(Titanic)
class(Titanic) <- c("sds270", class(Titanic))
Titanic
```

::: {.callout-important}

##### Q9

Define your own `print()` method for objects having class [your GitHub username]. Then add your class name to an object's listing of classes and print out that object.

```{r}
# PUT ANSWER HERE
```

:::

### Defining a new generic
In lecture, we defined a method for an existing generic function. Now, we will define a new generic function.

```{r}
print_emoji <- function(x, ...) {
  UseMethod("print_emoji")
}
```


Since we haven’t defined any methods yet, the function won’t actually work! However, we can start by writing a default method that will simply call the existing `print()` generic.

```{r}
print_emoji.default <- function(x, ...) {
  print(x, ...)
}
```

Now our `print_emoji()` function will work.

```{r}
print_emoji(starwars)
```


::: {.callout-important}

##### Q10

In the default method for `print_emoji()`, can we use `NextMethod()` instead of `print()`? Why or why not?

PUT ANSWER HERE

:::

::: {.callout-important}

##### Q11

Write a print_emoji() method for data.frames.

```{r}
# PUT ANSWER HERE
```

:::

## Turn-In

Stage, commit, and push your template file to your SDS 270 repository.