---
title: "R Code from Chapter 3"
format: pdf
---

## The Four Functions

As prefaced in code from Chapter 1, when you are looking at solving probabilities for common distributions, base R (and packages) have a family of functions used for particular items. The four functions are as follows:

- `d____()`: used for point probabilities (i.e. $P(X = x)$)
- `p____()`: used for cumulative probabilities (i.e. $P(X \leq x)$)
- `q____()`: used for identifying quantiles (i.e. $x$ for which $P(X \leq x) = q$)
- `r____()`: used to generate observations from distribution (i.e. $X \sim$___())

The abbreviation for the distribution you are looking at goes inside the blanks. For discrete RVs, here are those abbreviations:

- binom: binomial distribution with parameters $size = n$ and $prob = p$
- geom: geometric distribution with parameters $prob = p$
- hyper: hypergeometric distribution with parameters $m$, $n$, and $k$

We will talk about the continuous RVs later, but they follow a similar format. Let's explore how to use these functions when we are working with $X \sim Bin(n = 50, p = 0.3)$. When using the functions, we want to specify the $n$ and $p$ parameters in the function as secondary arguments. 

Suppose we cared about the probability of 4 successes for this random variable $X$. Then, we can use the following R code to find this point probability:

```{r}
dbinom(x = 4, size = 50, prob = 0.3)
```

Verify that this answer is correct by doing the probability by-hand -- you should get the same result!

Now let's say we want to find a cumulative probability, such as the probability of having no more than 2 successes. Then, we can use the `pbinom` function as follows:

```{r}
pbinom(q = 2, size = 50, prob = 0.3)
```

Note that $x = q$ when going from the `dbinom()` to `pbinom()` function. Say we want to find the probability of having greater than 2 successes. Then, we can find this one of two ways:

```{r}
# Complement rule
1 - pbinom(q = 2, size = 50, prob = 0.3)

# Use of lower.tail = F
pbinom(q = 2, size = 50, prob = 0.3, lower.tail = F)
```

Functionally, they give you the same result. The `lower.tail = F` argument tells the function that you want to look at the upper tail probability instead of the lower tail probability. 

Lastly, if I want to find the probability between 2 and 4 successes, then I can just write code that finds both $P(X \leq 4)$ and $P(X \leq 2)$, and then takes a difference of these two. Hence, we get:

```{r}
pbinom(q = 4, size = 50, prob = 0.3) - pbinom(q = 1, size = 50, prob = 0.3)
```

If we have an inclusive lower bound, we will need to lower the $x$ by 1. This is because we are removing the probability of having some value of $x$ for values below 2 in order to get our probability. If we have $q = 2$, we will only get the probability of $x = 3, 4$. 

**Note:** Watch out for whether values are included or excluded in the inequalities. If excluded, you will need to adjust your value of $x$. 

- Upper bound is exclusive: reduce $x$ by 1
- Lower bound is exclusive: keep $x$ at $a$

We may also be curious about the value of $x$ in which the percentage of the distribution beneath that value is a proportion of $p$. Here, we can use the `q___()` to find this value of $x$ given some percentile $p$ (i.e. $P(X \leq x) = p$). For example, suppose we wanted to find the number of successes for which 50\% of the distribution is beneath it. Then, we can use the following code:

```{r}
qbinom(p = 0.5, size = 50, prob = 0.3)
```
The $x$ value it returns is the first value of $x$ for which $P(X \leq x) \geq p$. This can be verified if we run:

```{r}
pbinom(q = 14, size = 50, prob = 0.3)
pbinom(q = 15, size = 50, prob = 0.3)
```
Lastly, if we want to generate/simulate $n$ observations from a distribution, we can use the `r___()` function to do so. As such, if we want to generate 100 observations from the $Bin(n = 50, p = 0.3)$ distribution, we can write this code:

```{r}
rbinom(n = 100, size = 50, p = 0.3)
```

Note that the $n$ argument that R uses here is different from the $n$ parameter used universally in all probability problems involving a binomial distribution. Nonetheless, if we run it, we get 100 observations that come from the specified binomial distribution. 

Generating and simulating $n$ observations can also come from any discrete RV with finite values for the support -- we just need the values of the support, their probabilities, and the `sample()` function. Suppose we evaluate the number of days someone runs in a week, with $x \in \{0, 1, \dots, 7\}$ and some $P(X = x)$ in which the sum of probabilities for values in the support add up to 1. Hence, we can do the following:

1. Define values of support using a vector we assign to $x$
2. Define probabilities for each value of support using a vector we assign to $p$
3. Use `sample()` function on $x$ and $p$, with a size indicating the number of observations we want and replace being set to true.

```{r}
x = c(0, 1, 2, 3, 4, 5, 6, 7)
p = c(0.0, 0.01, 0.09, 0.2, 0.25, 0.3, 0.13, 0.02)

sample(x, size = 100, replace = T, prob = p)
```
Again, we get observations from this specific random variable.
