---
title: "EDA: Part I"
format: html
---

## Libraries

```{r}
library(dplyr)
library(ggplot2)
library(gridExtra)
```


## Loading in Data

We can use the `read.csv()` function in order to import a dataset from our laptop. We need the following steps however:

1. Gather data and place it in an organized place in your computer storage.
  - Should be a .csv file
  - e.g. same folder as where your .qmd file will be in
2. Create .qmd/.R file
  - e.g. You can place file in same folder that has the data you're working with
3. Recall the file path to the .csv file
  - Relative: based on the file location of R file (you just need filename.csv)
  - Absolute: whole file path on your computer (e.g. C:/Users/.../filename.csv)
4. Use `read.csv("filepath")` and assign it to an object name of your choice
5. Run command and you should see it stored in your R environment

```{r}
movies <- read.csv("../datasets/movies.csv")
```

## Commands for Dataset Characteristics

```{r}
dim(movies) # number of rows and columns
nrow(movies) # number of rows
ncol(movies) # number of columns
length(movies$duration) # number of observations in column/vector
typeof(movies$duration) # R type of variable
```

## Frequency Tables

```{r}
table(movies$content_rating)
round(table(movies$content_rating) / nrow(movies), 4)
```

## Bar Plots

```{r}
# Creating a frequency bar plot
p1 <- ggplot(movies, mapping=aes(x=content_rating)) + 
  geom_bar(fill="brown") +
  ylab("Frequency") + 
  xlab("MPAA Rating")

# Creating a relative frequency bar plot
p2 <- ggplot(movies, mapping=aes(x=content_rating, y=(..count..)/sum(..count..))) +
  geom_bar(fill="light green") + 
  ylab("Relative Frequency") + 
  xlab("MPAA Rating")

grid.arrange(p1, p2, ncol=2)
```

## Pie and Waffle Charts

```{r}
library(baffle)

movies <- movies |> 
  mutate(content_rating = factor(content_rating))

factor(movies$content_rating, ordered = F, levels = unique(movies$content_rating))
```


```{r}
content_sum <- movies |> 
  group_by(content_rating) |> 
  summarise(count = n())

p3 <- ggplot(content_sum, aes(x = "", y = count, fill = content_rating)) + 
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  theme_void() +
  labs(fill = "MPAA Rating")

waffle(table(movies$content_rating))

```

## Histogram

```{r}
ggplot(movies, aes(x = duration)) +
  geom_histogram(bins = 40, color = "black", fill = "lightblue") +
  theme_bw() +
  xlab("Duration (mins)") +
  ylab("Frequency")
```

## Density Plot

```{r}
ggplot(movies, aes(x = duration)) +
  geom_density(color = "black", fill = "lightblue") +
  theme_bw() +
  xlab("Duration (mins)") +
  ylab("Rel. Frequency")
```

## Box Plot

```{r}
ggplot(movies, aes(x = duration)) +
  geom_boxplot(color = "black", fill = "lightblue") +
  theme_bw() +
  xlab("Duration (mins)")
```

