---
title: "Week 6, Lecture 1: Data Wrangling"
subtitle: "SDS 192: Introduction to Data Science"
author: "NAME HERE"
---

## Introduction

Use the following code below to load in the `midwest` data. Answer the following wrangling questions below:

```{r}
library(tidyverse)
data(midwest)
```

## Questions

::: {.callout-important}
###### Question 1

Using the help documentation for `midwest`, identify variables related to the following:

- county name
- state to which the county belongs to
- percent of population with known poverty status
- percent with professional degree
- whether county is in metropolitan area

Once done, select those columns using `dplyr` functions.

```{r}
# PUT SOLUTION HERE
```
:::

::: {.callout-important}
###### Question 2

With the columns selected from Q1, arrange the dataset based on the percent of populace with a professional degree in descending order, followed by percent of populace with known poverty status in ascending order.

```{r}
# PUT SOLUTION HERE
```

:::

::: {.callout-important}
###### Question 3

Adding to what was created in Q1, keep counties that are not in a metropolitan area and have less than 5\% of the populace with a professional degree.

```{r}
# PUT SOLUTION HERE
```

:::

::: {.callout-important}
###### Question 4

Tack a new column named `popProf` that contains the number of people who have a professional degree. Then, show only the county, state, and new column.

Hint: You will need two variables from the dataset to do this operation.

```{r}
# PUT SOLUTION HERE
```

:::

::: {.callout-important}
###### Question 5

Create a table that shows the following metrics in each state:

- Number of counties
- Total population
- Median percent with known poverty status
- Proportion of counties that are in metro area

```{r}
# PUT SOLUTION HERE
```

:::

## Turn-in

Save the .qmd file inside your activities repository. Please stage, commit, and push to GitHub.